---
title: "'Closest vaccine site' could favor those near Gillette vs MetroWest and Boston"
author: "Sharon Machlis"
date: "3/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
pacman::p_load(kableExtra, leaflet, DT, dplyr, htmltools, htmlwidgets, readxl, sf)
# Script to create the .Rds files is in distance.R
adult_population <- readRDS("adult_population.Rds")
site_map <- readRDS("vaccine_site_map.Rds")
sites <- readxl::read_xlsx("geocoded_sites.xlsx")



```

The Baker Administration's new centralized Covid-19 vaccine pre-registration system will dole out appointments to eligible resident based on "the availability of appointments at mass vaccination sites nearby," according to [the state's website](https://www.mass.gov/news/baker-polito-administration-announces-preregistration-system-for-mass-vaccination-locations-to). But how the system defines "nearby" will make an enormous difference in whether the system is fair.

"Residents will only be matched with the site that’s closest to them," the [Boston Globe reported on March 10](https://www.bostonglobe.com/2021/03/10/nation/what-know-about-states-new-preregistration-system-covid-19-vaccine-appointments/). If true, there would be vast discrepancies between a site like Gillette Stadium, which is closest to approximately 450,000 adults age 18+, and the Natick Mall, which is closest to more than 900,000. 

Under such rules, the state would have to significantly alter vaccine distributions to make sure there were similar doses _per capita_ and not per site. Otherwise, those living near sites like Gillette and the Doubletree Hotel in Danvers would have much greater access to appointments than residents of MetroWest and Boston. Or, the system would need to offer Gillette/Danvers/Dartmouth appointments to people who live closer to sites in Boston or Natick.

```{r}
adult_population %>%
  dplyr::arrange(AdultPopulation) %>%
  dplyr::rename(`Closest Site` = ClosestSite, `Adult Pop` = AdultPopulation) %>%
  kbl(digits = 3, format.args = list(big.mark = ",", 
  scientific = FALSE)) %>%
  kable_paper("hover", full_width = F)

```


Gillette had received considerably more vaccine than the Natick Mall in recent weeks, despite having the state's smallest nearby adult population. However, under the prior system, at least some residents in MetroWest, Boston, and elsewhere could choose to travel to Gillette. 

The new vaccine pre-registration process did not allow people to say which sites they'd be willing to travel to. It's unclear if/how many people in metro Boston outside the "closest-to-Gillette" zone will be offered appointments at Gillette or Danvers.

Below is a map of closest mass vaccination site for each city and town in Massachusetts.


```{r warning = FALSE, message=FALSE, echo=FALSE, fig.width=10}
ma_towns <- readRDS("ma_towns_for_map.Rds")
palette_colors <- RColorBrewer::brewer.pal(6, "Dark2")

pal <- colorFactor(palette_colors, ma_towns$ClosestSite)
ma_towns <- st_transform(ma_towns, "WGS84")

popup_text <- paste0("<strong>",
  ma_towns$Place, ":</strong><br />Est. adult population: ",
  scales::comma(ma_towns$estimate, accuracy = 1), " (margin of error ",
  scales::comma(ma_towns$moe, accuracy = 1), ")<br />Closest site: ",
  ma_towns$ClosestSite, "<br />Est. driving distance: ",
  ma_towns$Miles, " miles"
)  %>%   
  lapply(htmltools::HTML)

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(
    data = ma_towns,
    fillColor = ~pal(ma_towns$ClosestSite),
    fillOpacity = 0.9,
    weight = 0.2,
    smoothFactor = 0.2,
    label = popup_text
  ) %>%
  addLegend(pal = pal, 
            values = ma_towns$ClosestSite, 
            position = "bottomleft", 
            title = "Closest Mass Vaccine Site") %>%
  addMarkers(
    data = sites,
    ~Longitude,
    ~Latitude,
    label = sites$Site
  )

```



Those are rough estimates based on an overall city or town, not individual addresses, calculated by the [Google Maps Distance Matrix API](https://developers.google.com/maps/documentation/distance-matrix/overview). I suspect the state's system may measure an individual address (or Zip Code) and not simply the town where they live. This would further refine results and probably split a few communities as to which site is closer to various neighborhoods, but it seems unlikely this would dramatically shift the number of adults closest to each site. 

Note that because it was difficult to determine who in the immediate Boston area is closer to the Hynes Convention Center vs. the Reggie Lewis Center, I combined both sites into a single Boston total. (Some appointments at the Reggie Lewis Center are reserved for the immediate neighborhood.) I'm also not sure how the Google API calculates driving distances for Nantucket and Martha's Vineyard. 

Adult population 18+ data come from the [U.S. Census Bureau's 2018 American Community Survey](https://www.census.gov/acs/www/data/data-tables-and-tools/data-profiles/2018/).

More details by community:


```{r warning=FALSE, message=FALSE, echo=FALSE}
ma_towns_no_geo <- ma_towns
ma_towns_no_geo$geometry <- NULL
ma_towns_no_geo <- ma_towns_no_geo %>%
  select(Place, `Closest Site` = ClosestSite, Miles, `Est. Adult Pop` = estimate, `Margin of Error` = moe ) %>%
  arrange(Place)
DT::datatable(ma_towns_no_geo, filter = 'top', rownames = FALSE, options = list(
pageLength = 25,
lengthMenu = c(25, 50, 100)
)) %>%
  formatCurrency(4:5, digits = 0, currency = "")



```

You can see code for this analysis at [https://github.com/smach/vaccineSites](https://github.com/smach/distance). The code was written in R and used a number of packages including [tidycensus](https://walker-data.com/tidycensus/) to get population data, [leaflet](http://rstudio.github.io/leaflet/legends.html) for mapping, and [gmapsdistance](https://github.com/rodazuero/gmapsdistance) to access the Google Maps API for distance calculations.

Those familiar with R: The R Markdown document generating this page is at [https://github.com/smach/vaccineSites/blob/master/vaccine_distance.Rmd](https://github.com/smach/vaccineSites/blob/master/vaccine_distance.Rmd). The R script generating some of the data analysis for the R Markdown document is at [https://github.com/smach/vaccineSites/blob/master/distance.R](https://github.com/smach/vaccineSites/blob/master/distance.R).














